On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-10-16 Manuscript/manuscript.tex
few comments and edits

Figure~\ref{fig:RankScatterplots} shows the data that will be analyzed
with the rank-based methods (described in Section~\ref{sec:rank} with results in 
Section~\ref{subsec:Results_Rank}). In each of the three graphs, the reaction 
times for athletes who competed at the 2022 World Championships
are compared to their reaction times at another at which they competed. 
It is worth noting that each instance of the 2022 data is different, because
those who competed at both the 2019 and 2022 World Championships will be
different athletes than those who competed at both the 2022 and 2023 World
Championships. This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022.
The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. It
is also worth noting is that the fastest reaction time in the 2022 World
Championships Finals and Semifinals both came from Devon Allen, but the 
difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified.

@@

The data under consideration is summarized using side-by-side boxplots in 
Figure~\ref{fig:Boxplot}. It is clear that the reaction times in the 2022
box plot are lower (indicating faster reaction times) than in many of the other
years. The median reaction time in 2022 was 0.129 seconds,  compared to 0.156
seconds, which is what \citet{brosnan2017effects} found when looking at data
from 1999 to 2014. We can also see from Figure~\ref{fig:Boxplot} that the 
reaction times clearly differ from year to year. The venue of the World 
Championships changes each year, so weather or climate related factors such as 
humidity, precipitation, elevation, may indeed play a role, but additionally,
both the technology and false start penalization rules have changed over the 
study period.

@@

Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$v_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $h_{i/j}$ 
is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

Table~\ref{tab:Clusrankresults} shows the results generated from both the 
permutation and asymptotic rank-based tests for the three comparisons, 
as discussed previously in Section~\ref{sec:rank}. The national versus international
comparison produced lower p-values than the 2019 versus 2022 comparison, with the
national vs international permutation test producing a p-value of $2\cdot10^{-6}$.
This is a highly significant test result and shows that athletes who competed at the
2022 World Track and Field Championships were on average, significantly faster
at reacting than they were just a few months prior. For the 2019 versus 2022
same athlete comparison with permutation test, the p-value was 
$0.003$, which is significant at $\alpha = 0.01$. The result of
comparing 2022 and 2023 was very similar (but not exactly the same) to the above
result, as the p-value from the permutation test was $0.003$. Taken together, these three
tests help show that athletes competing at the 2022 World Track and Field
Championships had faster reaction times than when they had competed previously.

@@

The estimated parameters of the GLMM model under the log link are shown in 
Table~\ref{tab:Gamma_parameters}, with both including and excluding 2022 data.
Three models were considered, differing in there random effects: venue effect
only; heat effect only, and both venue and heat effects. Reported are the AIC
and estimated model parameters: the standard deviation
of the heat effect $\sigma_h$, the standard deviation of the venue
effect~$\sigma_v$, the intercept~$\alpha$, and the
square root of the dispersion parameter $\phi$. With or without 2022,
the GLMM with both venue and heat random effects was the best model as evidenced
by the AIC compared to the other models with only one random
effect.  When we exclude 2022 from the full model, we observe small but intriguing
discrepancies in the models. The standard deviation of the venue effect, 
$\sigma_v$, increases from 0.33 to 0.40 when 2022 is included, however, the
changes to the estimates of other parameters
remained relatively similar, such as the standard deviation of the heat effect
and the dispersion parameter.

@@

Figure~\ref{fig:VenueEffects} further illustrates the venue effect from 1999 to
2023 based on the full GLMM both with and without 2022.  The temporal patterns
of the two sets of effects look very similar, except that the positive effects
estimated with 2022 included have higher magnitude than those estimated without
2022, which compensates the negative effect of the 2022 with a large magnitude.
Both plots agree to some extent with the pattern of the centers of the box plots in
Figure~\ref{fig:Boxplot}. The mismatch is due to the venue effect, which has a
higher standard deviation than the venue effect. The venue effect for 2022 is
lower than that for any other years, but may not be as extreme as one would
think considering the history of reaction times going back to 1999.

@@

The GLMM with both venue and heat random effects helps us to assess how extreme
a reaction time below a given threshold is. The probability of observing a reaction
time below a threshold, assuming no intentional false starts, can be
approximated by generating a large number of realizations from the fitted
GLMM. We used 10 millions realizations in the analysis. 
Table~\ref{tab:Sim_probability} summarizes the estimated probabilities of
observing a reaction time below 0.08, 0.09, and 0.10 seconds under two different
scenarios: one excluding and the other including data from 2022. Apparently,
excluding 2022 lowered the probability of observing a fast reaction time, but
the difference is not in a any big magtinude. For example, the probability for
seeing a reaction time below 0.10 seconds changes from $3.54\cdot 10^{-3}$ to
$4.20\cdot 10^{-3}$ when 2022 is included. They corresponding to roughly one in
every 230 starts and 283 starts, respectively. The table shows that changing the
reaction time barrier from 0.1 seconds to 0.08 seconds drastically reduces the
chance of observing a reaction time below the barrier, from one in roughly every
283 starts to one in every 29851 starts using data with 2022 included. The
results substantiates the recommendations by \citet{komi2009iaaf}.

>>>>

Figure~\ref{fig:RankScatterplots} shows the data that will be analyzed
with the rank-based methods (described in Section~\ref{sec:rank} with results in 
Section~\ref{subsec:Results_Rank}). In each of the three graphs, the reaction 
times for athletes who competed at the 2022 World Championships
are compared to their reaction times at another at which they competed. 
It is worth noting that each instance of the 2022 data is different, because
those who competed at both the 2019 and 2022 World Championships will be
different athletes than those who competed at both the 2022 and 2023 World
Championships. This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022.
%The 2022 World
%Championship Heats and Semifinals both had on average lower reaction times than
%their respective 2019 races, and the 2022 finals were substantially faster. 
The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective other (2019, 2023 or 2022 national) races, and the 2022 World 
Championship finals were substantially faster. 
\eds{please check change}
It
is also worth noting is that the fastest reaction time in the 2022 World
Championships Finals and Semifinals both came from Devon Allen, but the 
difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified.

@@

The data under consideration is summarized using side-by-side boxplots in 
Figure~\ref{fig:Boxplot}. It is clear that the reaction times in the 2022
boxplot are lower (indicating faster reaction times) than in many of the other
years. The median reaction time in 2022 was 0.129 seconds,  compared to 0.156
seconds, which is what \citet{brosnan2017effects} found when looking at data
from 1999 to 2014. We can also see from Figure~\ref{fig:Boxplot} that the 
reaction times clearly differ from year to year. The venue of the World 
Championships changes each year, so weather or climate related factors such as 
humidity, precipitation, elevation, may indeed play a role, but additionally,
both the technology and false start penalization rules have changed over the 
study period.

@@

Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function, $\alpha$ is a (fixed) intercept,
$v_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $h_{i/j}$ 
is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

Table~\ref{tab:Clusrankresults} shows the results generated from both the 
permutation and asymptotic rank-based tests for the three comparisons, 
as discussed previously in Section~\ref{sec:rank}. The national versus international
comparison produced lower p-values than the 2019 versus 2022 comparison, with the
national vs international permutation test producing a p-value of $2\cdot10^{-6}$.
This is a highly significant test result and shows that athletes who competed at the
2022 World Track and Field Championships were on average, significantly faster
at reacting than they were just a few months prior. For the 2019 versus 2022
same athlete comparison with permutation test, the p-value was 
$0.003$, which is significant at $\alpha = 0.01$. 
\eds{According to the table, this p-value is 0.028 (significant at the 0.05 
level.)}
The result of
comparing 2022 and 2023 was very similar (but not exactly the same) to the above
result, as the p-value from the permutation test was $0.003$.
\eds{According to the table, this p-value is $0.011$.} 
Taken together, these three
tests help show that athletes competing at the 2022 World Track and Field
Championships had faster reaction times than when they had competed at other
competitions.

@@

The estimated parameters of the GLMM model under the log link are shown in 
Table~\ref{tab:Gamma_parameters}, both excluding and including 2022 data.
Three models were considered, differing in their random effects: venue effect
only; heat effect only, and both venue and heat effects. Reported are the AIC
and estimated model parameters: the standard deviation
of the heat effect $\sigma_h$, the standard deviation of the venue
effect~$\sigma_v$, the intercept~$\alpha$, and the
square root of the dispersion parameter $\phi$. With or without 2022,
the GLMM with both venue and heat random effects was the best model as evidenced
by the AIC compared to the other models with only one random
effect.  When we exclude 2022 from the full model, we observe small but intriguing
discrepancies in the models. The standard deviation of the venue effect, 
$\sigma_v$, increases from 0.33 to 0.40 when 2022 is included, however, the
 estimates of other parameters
remained relatively similar, such as the standard deviation of the heat effect
and the dispersion parameter.

@@

Figure~\ref{fig:VenueEffects} further illustrates the venue effect from 1999 to
2023 based on the full GLMM both with and without 2022.  The temporal patterns
of the two sets of effects look very similar, except that the positive effects
estimated with 2022 included have higher magnitude than those estimated without
2022, which compensates the negative effect of the 2022 with a large magnitude.
Both plots agree to some extent with the pattern of the centers of the boxplots in
Figure~\ref{fig:Boxplot}. The mismatch is due to the venue effect, which has a
higher standard deviation than the venue effect. 
\eds{I do not understand what is meant by the previous sentence.}
The venue effect for 2022 is
lower than that for any other year, but may not be as extreme as one would
think considering the history of reaction times going back to 1999.

@@

The GLMM with both venue and heat random effects helps us to assess how extreme
a reaction time below a given threshold is. The probability of observing a reaction
time below a threshold, assuming no intentional false starts, can be
approximated by generating a large number of realizations from the fitted
GLMM. We used 10 millions realizations in the analysis. 
Table~\ref{tab:Sim_probability} summarizes the estimated probabilities of
observing a reaction time below 0.08, 0.09, and 0.10 seconds under two different
scenarios: one excluding and the other including data from 2022. Apparently,
excluding 2022 lowered the probability of observing a fast reaction time, but
the difference is not big in magnitude. For example, the probability for
seeing a reaction time below 0.10 seconds changes from $3.54\cdot 10^{-3}$ to
$4.20\cdot 10^{-3}$ when 2022 is included. These values correspond to roughly one in
every 230 starts and 283 starts, respectively. The table shows that changing the
reaction time barrier from 0.1 seconds to 0.08 seconds drastically reduces the
chance of observing a reaction time below the barrier, from one in roughly every
283 starts to one in every 29851 starts using data with 2022 included. The
results substantiates the recommendations by \citet{komi2009iaaf}.
